Clustering Constraints on the Relative Sizes of Central and Satellite Galaxies

### 2017-11-08 paper/galsize_paper.tex
edited S 3 and 4

\section{Introduction}
\label{sec:intro}

@@

\section{Data and Simulations}
\label{sec:data}

Using galaxies observed in the Sloan Digital Sky Survey (SDSS) with the stellar mass measurements taken from the MPA-JHU catalog \citep{kauffmann_etal03,brinchmann_etal04}, we define volume-limited samples of galaxies according to the same $\mstar-$completeness cuts used in \citet{behroozi_etal15} (see Figure 2). We supplement the MPA-JHU catalog with measurements of half-light radius, $\rhalf,$ derived from galaxy profile decompositions provided by \citet{meert_etal15}. The \citet{meert_etal15} catalog is based on Data Release 10 of SDSS \citep{ahn_etal14}, with improvements to the photometry pipeline and light profile fitting methods \citep{vikram_etal10,bernardi_etal13,bernardi_etal14,meert_etal13}. In the version of this catalog that we use,\footnote{Our $\mstar$ and $\rhalf$ measurements are derived from different photometry pipelines; this is driven by our desire for consistency with the \citet{behroozi_etal15} completeness cuts used in our clustering measurements. We note that we have repeated our analysis for stellar mass measurements based on \citet{meert_etal15}, finding only minor quantitative, and no qualitative changes to our results.} two-dimensional $r-$band profiles were fit with a two-component de Vaucouleurs + exponential profile to determine the half-light radius of total $r-$band luminosity, $\rhalf.$

@@

Using this technique, for any $\mstar-$threshold sample, the stellar mass function of the ``large" and ``small" subsamples are identical. We illustrate this for the particular case of $\mstar>10^{10.25}\msun$ in left panel of Figure~\ref{fig:sizedefinition}, which shows histograms of stellar mass for the two subsamples. The right panel of Figure~\ref{fig:sizedefinition} compares histograms of the two size distributions, which partially overlap due to the variation in $\median{\rhalf}{\mstar}$ across the $\mstar-$ range of the threshold sample.

%---------------------------------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=8cm]{FIGS/sdss_small_large_sample_definitions.pdf}
\caption{
{\bf Definition of ``small" and ``large" galaxies.} For a volume-limited SDSS galaxy sample defined by $M_{\ast}>10^{10.25}M_{\odot},$ we visually demonstrate how we classify galaxies into ``small" and ``large" subsamples. As described in detail in \S\ref{subsec:sizedef}, we compute the median value $\median{\rhalf}{\mstar}$ at the stellar mass of each galaxy in the sample, and divide the sample into two based on this $\mstar-$dependent cut. The {\em left panel} compares the stellar mass functions $\phi(\mstar)$ of the two samples, confirming that our method for separating ``small" from ``large" galaxies yields subsamples with statistically indistinguishable $\phi(\mstar)$. The {\em right panel} shows histograms of $\rhalf$ for the two subsamples, which partially overlap due to the finite range of $\mstar$ in the volume-limited sample. 
}
\label{fig:sizedefinition}
\end{figure}
%-----------------------------------------------------------------------------------------------------

@@

In one class of models we explore, we suppose that stellar mass $\mstar$ is the statistical regulator of $\rhalf,$ so that galaxy sizes are drawn from a log-normal distribution with $0.2$ dex of scatter centered at $\median{\rhalf}{\mstar}.$ To implement this model, for simplicity we tabulate $\median{\rhalf}{\mstar}$ directly from the data, rather than pursue a parametric form \citep[see, e.g.,][]{zhang_yang17}.

@@

Motivated by \citet{kravtsov13}, we additionally explore a model in which $\rhalf$ is linearly proportional to halo virial radius:\footnote{We note that the normalization factor in \citet{kravtsov13} is 0.0125, where as our normalization is 0.01. The difference derives from the choice of halo radius definition: \citet{bryan_norman98} in the present work, compared to the $200c$ definition used in \citet{kravtsov13}.} 
\beq
\label{eq:fiducial_model}
\rhalf = 0.01\rvir.
\eeq
For the virial radius of halos and subhalos, we use $\rmpeak:$ the value of $\rvir$ in physical units of $\kpc$ measured at the time of peak subhalo mass. The relationship between $\rmpeak$ and $\mpeak$ is given by 
\beq
\mpeak\equiv\frac{4\pi}{3}\rmpeak^{3}\Delta_{\rm vir}(\zpeak)\rho_{\rm m}(\zpeak),
\eeq
where for $\Delta_{\rm vir}(\zpeak)$ we use the fitting function to the ``virial" definition used in \citet{bryan_norman98}. For the model we refer to as the ``$\rvir-$based model", we add uncorrelated log-normal scatter of $\sigmarhalf=0.2$ dex to generate a Monto Carlo realization of the model population.

\section{Results}
\label{sec:results}

\subsection{Size-Mass Scaling Relation}
\label{subsec:one_point_function}

In Figure \ref{fig:scatter_plot} we show the scaling of galaxy size $\rhalf$ with $\mstar.$ The black curve enveloped by the gray bands shows the scaling relation for SDSS galaxies, while the blue curve shows the median relation  $\median{\rhalf}{\mstar}$ implied by the $\rvir-$based model described in \S\ref{subsubsec:rvirmodel}. This figure shows that models in which $\rhalf\propto\rvir$ can naturally give rise to the characteristic curvature in the $\rhalf-\mstar$ relation, confirming the results in \citet{kravtsov13} in a forward modeling context.

\subsection{Size-Dependent Clustering Measurements}
\label{subsec:clustering_results}

In Figure~\ref{fig:rvir_only_clustering_absolute}, we present new measurements of the $\rhalf-$dependence of projected galaxy clustering, $\wproj(\rproj).$ We measure $\wproj(\rproj)$ separately for large and small subsamples for four different $\mstar$ thresholds, $\mstar>10^{9.75}\msun,$ $\mstar>10^{10.25}\msun,$ $\mstar>10^{10.75}\msun,$ and $\mstar>10^{11}\msun.$ For each threshold, we split the galaxies into ``large" and ``small" subsamples; as described in \S\ref{subsec:sizedef} and illustrated in Figure \ref{fig:sizedefinition}, our size-based selection is defined so that the subsamples have identical stellar mass functions. Red points with jackknife-estimated error bars show SDSS measurements of $\wproj(\rproj)$ for small galaxy samples, blue points show the same for large galaxies. Solid curves show $\wproj$ as predicted by the $\rvir-$based model described in \S\ref{subsubsec:rvirmodel}.

The salient feature of these clustering measurements is that small galaxies cluster more strongly than large galaxies of the same stellar mass. This feature also holds true for galaxies predicted by the $\rvir-$based model. This result may be surprising, since $\rhalf\propto\rvir,$ halo mass $\rvir\propto\mhalo^{1/3},$ and clustering strength increases with $\mvir.$ Based on this simple argument, one would expect that large galaxies would be the more strongly clustered. We provide a resolution to this conundrum in \S\ref{subsec:censat_sizes}; before doing so, we first examine the clustering test of the $\rvir-$based model in more detail.

@@

The points and curves in Figure \ref{fig:clustering_ratio_upshot} are all negative: small galaxies cluster more strongly relative to large. This presentation of the measurement makes plain that the underlying signal of $\rhalf-$dependent clustering is strongest for samples with smaller stellar mass; as $\mstar$ increases, the signal weakens and nearly vanishes for $\mstar\gtrsim10^{11}\msun.$ Strikingly, the $\rvir-$based model exhibits this $\mstar-$dependent behavior, as well as the scale-dependence of the observed clustering signal at each $\mstar.$ As described in \S\ref{subsec:censat_sizes} below, we attribute the success of this prediction to the relative sizes of central vs. satellite galaxies.

\subsection{Testing the Impact of Satellite Mass Loss}
\label{subsec:mstar_stripping}
%---------------------------------------------------------------------------------------------------
\begin{figure*}
\centering
\includegraphics[width=12cm]{FIGS/alt_model_wp_ratios.pdf}
\caption{
{\bf The subdominant role of tidal stripping.}
In all panels, the axes and points with error bars are the same as in Figure \ref{fig:clustering_ratio_upshot}. The solid green curves shows the prediction of a model where $\rhalf$ is statistically set by $\mstar,$ in the complete absence of satellite mass loss. Such a model would predict zero $\rhalf-$dependence to galaxy clustering, in gross tension with observations. The solid purple curves show results for a model in which satellites lose mass after infall in a manner similar to what is seen in high-resolution hydrodynamical simulations, as described in \S\ref{subsubsec:strippingmodel}. This produces satellites that are smaller than centrals, but the effect is too mild to correctly capture the observed clustering. Evidently, satellite-specific mass stripping plays a sub-dominant role in setting the relative size of centrals and satellites.
}
\label{fig:mstarmodelclustering}
\end{figure*}
%-----------------------------------------------------------------------------------------------------

The green curves in Figure~\ref{fig:mstarmodelclustering} test the clustering predictions of the model described in \S\ref{subsubsec:strippingmodel} {\em before} applying the satellite mass loss prescription. Thus the green curves pertain to a model where $\rhalf$ follows a purely random log-normal distribution centered at $\median{\rhalf}{\mstar}.$ By construction, this model predicts no dependence at all of galaxy clustering upon $\rhalf$ at fixed $\mstar,$ simply because the scatter of $\rhalf$ about $\median{\rhalf}{\mstar}$ is uncorrelated with any other variable. The application of satellite mass loss introduces correlations in the scatter: after mass loss, satellites have smaller sizes than centrals of the same $\mstar$ due to post-infall stripping.

The purple curves in Figure \ref{fig:mstarmodelclustering} show that satellite mass loss impacts $\rhalf-$dependent clustering in the expected manner. The ``small" galaxy population has a higher satellite fraction due to satellites being smaller than centrals of the same $\mstar,$ and so in this model small galaxies cluster more strongly relative to large. However, the magnitude of the effect is not strong enough to produce clustering predictions that are consistent with SDSS. Evidently it is difficult to strip enough mass from satellites so that the $\rhalf-$dependent clustering predictions are in agreement with observations. As we argue in \S\ref{sec:interpretation}, this supports the conclusion that the relative difference between central and satellite size is at least partially in place at the time of satellite infall.

\subsection{The Role of Galaxy Color}
\label{subsec:colormorph}

In this section we take a preliminary look at how galaxy clustering exhibits simultaneous dependence upon broadband color and size. Beginning from the $\mstar>10^{10.25}\msun$ sample used above, we first divide the sample into ``red" and ``blue" subsamples according to a $g-r=0.65$ cut, the rough location of the trough of the green valley for this stellar mass. For each color-selected subsample, we separately measure $\median{\rhalf}{\mstar; {\rm red}}$ and $\median{\rhalf}{\mstar; {\rm blue}},$ and use these median size values to split each subsample into two.

Figure \ref{fig:colorclustering} shows the clustering of large vs. small galaxies, separately for red and blue samples. For each color-selected sample, large galaxies are slightly more strongly clustered relative to small galaxies. Comparing Figure \ref{fig:colorclustering} to the the upper right panel of Figure \ref{fig:rvir_only_clustering_absolute} shows the dramatic impact of galaxy color selection upon $\rhalf-$dependent clustering. For $\mstar-$complete samples, small galaxies cluster much stronger than large galaxies; for color-selected samples, the reverse is true, and the magnitude of the effect weakens considerably. We discuss a halo model interpretation of these trends in \S\ref{sec:interpretation}.

>>>>

%----------------------------
\section{Introduction}
\label{sec:intro}
%----------------------------

@@

%---------------------------------------------------------------------------------------------------
\begin{figure*}
\centering
\includegraphics[width=\textwidth]{FIGS/sdss_small_large_sample_definitions.pdf}
\caption{
{\bf Definition of ``small" and ``large" galaxies.} For a volume-limited SDSS galaxy sample defined by $M_{\ast}>10^{10.25}M_{\odot},$ we visually demonstrate how we classify galaxies into ``small" and ``large" subsamples. As described in detail in \S\ref{subsec:sizedef}, we compute the median value $\median{\rhalf}{\mstar}$ at the stellar mass of each galaxy in the sample, and divide the sample into two based on this $\mstar-$dependent cut. The {\em left panel} compares the stellar mass functions $\phi(\mstar)$ of the two samples, confirming that our method for separating ``small" from ``large" galaxies yields subsamples with statistically indistinguishable $\phi(\mstar)$. The {\em right panel} shows histograms of $\rhalf$ for the two subsamples, which partially overlap due to the finite range of $\mstar$ in the volume-limited sample.  \textcolor{blue}{AK: x axis label in the left panel should be just $M_\star\ (M_\odot)$ without $\log_{10}$. I think this figure should be either displayed full text width like this, or made into two vertical panels and displayed in a single column. Otherwise it is too small. I wonder also whether the panel titles at the top are  needed.}
}
\label{fig:sizedefinition}
\end{figure*}
%-----------------------------------------------------------------------------------------------------

%-----------------------------------------
\section{Data and Simulations}
\label{sec:data}
%-----------------------------------------

Using galaxies observed in the Sloan Digital Sky Survey (SDSS) with the stellar mass measurements taken from the MPA-JHU catalog \citep{kauffmann_etal03,brinchmann_etal04}, we define volume-limited samples of galaxies according to the same $\mstar-$completeness cuts used in \citet{behroozi_etal15} (see Figure 2). \textcolor{blue}{AK: which DR the catalog is taken from?} We supplement the MPA-JHU catalog with measurements of half-light radius, $\rhalf,$ derived from galaxy profile decompositions provided by \citet{meert_etal15}. The \citet{meert_etal15} catalog is based on Data Release 7 of SDSS \citep{abazajian_etal09}, with improvements to the photometry pipeline and light profile fitting method and, especially, in background subtraction \citep{vikram_etal10,bernardi_etal13,bernardi_etal14,meert_etal13}. In the version of this catalog that we use,\footnote{Our $\mstar$ and $\rhalf$ measurements are derived from different photometry pipelines; this is driven by our desire for consistency with the \citet{behroozi_etal15} completeness cuts used in our clustering measurements. We note that we have repeated our analysis for stellar mass measurements based on \citet{meert_etal15}, finding only minor quantitative, and no qualitative changes to our results.} two-dimensional $r-$band profiles were fit with a two-component de Vaucouleurs + exponential profile to determine the half-light radius of total $r-$band luminosity, $\rhalf.$

@@

Using this technique, for any $\mstar-$threshold sample, the stellar mass function of the ``large" and ``small" subsamples are identical. We illustrate this for the particular case of $\mstar>10^{10.25}\msun$ in left panel of Figure~\ref{fig:sizedefinition}, which shows stellar mass functions for the two subsamples. The right panel of Figure~\ref{fig:sizedefinition} compares histograms of the two size distributions of these subsamples, which partially overlap due to the variation in $\median{\rhalf}{\mstar}$ across the $\mstar-$ range of the threshold sample.

In addition to stellar mass galaxy clustering depends on galaxy color \citep[e.g.,][]{zehavi_etal11}. Thus, we will also consider 
dependence of clustering on galaxy size while controlling for color. The results for the subsamples split in stellar mass and color are presented in \S\ref{subsec:colormorph}.

@@

\textcolor{blue}{AK: I don't think the name of the model is descriptive here, but I could not yet come up with a better one.}
In the first model we explore, we assume that stellar mass $\mstar$ controls $\rhalf,$ so that galaxy sizes are drawn from a log-normal distribution with $0.2$ dex of scatter centered at $\median{\rhalf}{\mstar}.$ To implement this model, for simplicity we tabulate $\median{\rhalf}{\mstar}$ directly from the data, rather than pursue a parametric form \citep[see, e.g.,][]{zhang_yang17}. Note that in this model the theoretical sample of model reproduces the observed $\rhalf-\mstar$ relation by construction.

@@

As noted in \S\ref{sec:intro}, \citet{kravtsov13} showed that galaxy sizes are on average proportional to the virial radius of their host halo. Motivated by these results, we  explore a model in which $\rhalf$ scales linearly with the halo virial radius:\footnote{We note that the normalization factor in \citet{kravtsov13} is 0.015, where as our normalization is 0.01. The difference derives from the choice of halo radius definition: \citet{bryan_norman98} in the present work, compared to the $200c$ definition used in \citet{kravtsov13} and the fact that the latter study used deprojected (larger) sizes for spheroidal galaxies. } 
\beq
\label{eq:fiducial_model}
\rhalf = 0.01\rvir.
\eeq
For the virial radius of halos and subhalos, we use $\rmpeak:$ the value of $\rvir$ in physical units of $\kpc$ measured at the time when halo mass reached its maximum.  The relationship between $\rmpeak$ and $\mpeak$ is given by 
\beq
\mpeak\equiv\frac{4\pi}{3}\rmpeak^{3}\Delta_{\rm vir}(\zpeak)\rho_{\rm m}(\zpeak),
\eeq
where for $\Delta_{\rm vir}(\zpeak)$ we use the fitting function to the ``virial" definition used in \citet{bryan_norman98}. For the model we refer to as the ``$\rvir-$based model", we add uncorrelated log-normal scatter of $\sigmarhalf=0.2$ dex to generate a Monto Carlo realization of the model population.

%-----------------------
\section{Results}
\label{sec:results}
%-----------------------

%--------------------------------------------------------
\subsection{Size-Mass Scaling Relation}
\label{subsec:one_point_function}
%-------------------------------------------------------

In Figure \ref{fig:scatter_plot} we show the scaling of galaxy size $\rhalf$ with $\mstar.$ The black curve enveloped by the gray bands shows the scaling relation for SDSS galaxies, while the blue curve shows the median relation  $\median{\rhalf}{\mstar}$ predicted using the $\rvir-$based model described in \S\ref{subsubsec:rvirmodel}. This figure shows that models in which $\rhalf\propto\rvir$ can naturally match the observed $\rhalf-\mstar$ relation, confirming the results of \citet{kravtsov13} in a forward modeling context.

%---------------------------------------------------------------------------
\subsection{Size-Dependent Clustering Measurements}
\label{subsec:clustering_results}
%---------------------------------------------------------------------------

In Figure~\ref{fig:rvir_only_clustering_absolute}, we present new measurements of the $\rhalf-$dependence of projected galaxy clustering, $\wproj(\rproj).$ We measure $\wproj(\rproj)$ separately for large and small subsamples for four different $\mstar$ thresholds, $\mstar>10^{9.75}\msun,$ $\mstar>10^{10.25}\msun,$ $\mstar>10^{10.75}\msun,$ and $\mstar>10^{11}\msun.$ For each threshold, we split the galaxies into ``large" and ``small" subsamples, as described in \S\ref{subsec:sizedef}. As shown in Figure \ref{fig:sizedefinition}, our size-based samples  have similar stellar mass functions. Red points with jackknife-estimated error bars show SDSS measurements of $\wproj(\rproj)$ for samples of small galaxies, while blue points show the measured correlation functions for samples of large galaxies. Solid curves show $\wproj$ as predicted for these subsamples by the $\rvir-$based model of galaxy sizes described in \S\ref{subsubsec:rvirmodel}.

The salient feature of these clustering measurements is that small galaxies cluster more strongly than large galaxies of the same stellar mass in all but the largest $\mstar$ samples. Remarkably, both the trend and the quantitative difference in the shape of the correlation functions of small and large galaxy samples are reproduced by the $\rvir-$based model. This may be surprising, since $\rhalf\propto\rvir,$ halo mass $\rvir\propto\mhalo^{1/3},$ and clustering strength of halos increases with $\mvir.$ Based on this simple argument, one would expect that large galaxies would be the more strongly clustered. We provide a resolution to this conundrum in \S\ref{subsec:censat_sizes}. However,  before doing so, we first examine the clustering test of the $\rvir-$based model in more detail.

@@

The points and curves in Figure \ref{fig:clustering_ratio_upshot} are all negative: small galaxies cluster more strongly relative to large. This presentation of the measurement makes it clear that the underlying signal of $\rhalf-$dependent clustering is strongest for samples with smaller stellar mass; as $\mstar$ increases, the signal weakens and nearly vanishes for $\mstar\gtrsim10^{11}\msun.$ Strikingly, the $\rvir-$based model reproduces this $\mstar-$dependent behavior, as well as the scale-dependence of the observed clustering signal at each $\mstar.$ As described in \S\ref{subsec:censat_sizes} below, we attribute the success of this prediction to the systematic difference of sizes of central and satellite galaxies.

%--------------------------------------------------------------------------
\subsection{Testing the Impact of Satellite Mass Loss}
\label{subsec:mstar_stripping}
%--------------------------------------------------------------------------------------
\begin{figure*}
\centering
\includegraphics[width=12cm]{FIGS/alt_model_wp_ratios.pdf}
\caption{
{\bf The subdominant role of tidal stripping.}
In all panels, the axes and points with error bars are the same as in Figure \ref{fig:clustering_ratio_upshot}. The solid green curves shows the prediction of a model where $\rhalf$ is statistically set by $\mstar,$ in the complete absence of satellite mass loss. Such a model would predict zero $\rhalf-$dependence to galaxy clustering, in gross tension with observations. The solid purple curves show results for a model in which satellites lose mass after infall in a manner similar to what is seen in high-resolution hydrodynamical simulations, as described in \S\ref{subsubsec:strippingmodel}. This produces satellites that are smaller than centrals, but the effect is too mild to correctly capture the observed clustering. Evidently, satellite-specific mass stripping plays a sub-dominant role in setting the relative size of centrals and satellites.
}
\label{fig:mstarmodelclustering}
\end{figure*}
%-----------------------------------------------------------------------------------------------------

Figure~\ref{fig:mstarmodelclustering} shows a similar comparison of the observed clustering ratios to the model
in which size is controlled by $\mstar$ with sizes of satellites affected by stellar mass loss due to tides (\S\ref{subsubsec:strippingmodel}). Green solid lines show predictions of the model that does not account for stellar
mass loss of satellites due to tides. In this model $\rhalf$ follows a purely random log-normal distribution centered at $\median{\rhalf}{\mstar}$ with the median relation matched to the observed relation by construction.  This model thus predicts no dependence of galaxy clustering upon $\rhalf$ at fixed $\mstar,$ simply because the scatter of $\rhalf$ about $\median{\rhalf}{\mstar}$ is uncorrelated with any other variable. The figure shows that this model clearly is at odds with significant dependence of clustering on size exhibited by SDSS galaxies.

Inclusion of satellite mass loss in the model introduces correlations in the scatter: after mass loss, satellites have smaller sizes than centrals of the same $\mstar$ due to post-infall stripping.
The purple curves in Figure \ref{fig:mstarmodelclustering} show that satellite mass loss impacts $\rhalf-$dependent clustering in the expected manner. The ``small" galaxy population has a higher satellite fraction due to satellites being smaller than centrals of the same $\mstar,$ and so in this model small galaxies cluster more strongly relative to large. However, the magnitude of the effect is not sufficiently strong  to produce clustering predictions that are consistent with observations. Evidently it is difficult to strip enough mass from satellites so that the $\rhalf-$dependent clustering predictions are in agreement with observations. As we argue in \S\ref{sec:interpretation}, this supports the conclusion that the relative difference between central and satellite size is largely in place at the time of satellite infall.

%----------------------------------------------------
\subsection{The Role of Galaxy Color}
\label{subsec:colormorph}
%----------------------------------------------------

We now examine how galaxy clustering depends on size when we control both for stellar mass and color dependence of galaxy clustering. Using the $\mstar>10^{10.25}\msun$ sample used above, we first divide the sample into ``red" and ``blue" subsamples using $g-r=0.65$ as a boundary, which approximately corresponds to the trough of the green valley for this stellar mass. For each color-selected subsample, we separately measure $\median{\rhalf}{\mstar; {\rm red}}$ and $\median{\rhalf}{\mstar; {\rm blue}},$ and use these median size values to split each subsample into two.

Figure \ref{fig:colorclustering} shows the clustering of large vs. small galaxies, separately for red and blue samples. For each color-selected sample, large galaxies are slightly more strongly clustered relative to small galaxies. Comparing Figure \ref{fig:colorclustering} to the the upper right panel of Figure \ref{fig:rvir_only_clustering_absolute} shows the size dependence for color-selected samples is dramatically weaker.   For $\mstar-$complete samples, small galaxies cluster much stronger than large galaxies; for color-selected samples, the reverse is true, but the difference is much smaller and is confined to a limited range of scales.  Although the two results may seem surprising, even contradictory, they can, in fact be 
explained within a unified simple model, as will be discussed in \S\ref{sec:interpretation}.
